---
title: "Mind & Skin Redcap Database"
author: "Ella Jade Alexander"
format: html
editor: visual
---

# Redcap Database

## Setup

### Setting Base Directory

Assuming that this quarto is in the Redcap/scripts folder, we want to set it to the Redcap folder.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))

```

### Loading Necessary Libraries

```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```

### Loading Raw Data

```{r}

# Read Redcap csv with snake_case
redcap_raw <- read_csv("data/raw/MindSkinVer40_DATA_2025-01-22_1418.csv")

# Preview the raw database
redcap_raw
```

### Metadata: Data Dictionary

Here, we are creating a data dictionary that shows the human-readable, snake_case labels, and variable coding all in one place.

When loading the DATA_LABELS csv into R, the names of the columns are renamed for each to be unique, since there are multiple repeats. So we can't just load it in normally.

Also, for this to work, there have to be equal numbers of columns in both .csv files. The stopifnot function ensures that before continuing.

```{r}
# Extract clean column labels from the LABELS file
label_names <- names(read_csv("data/raw/MindSkinVer40_DATA_LABELS_2025-01-22_1419.csv", col_names = TRUE, n_max = 0))

# Read just the first line (header) of the LABELS CSV as a plain string
header_line <- read_lines("data/raw/MindSkinVer40_DATA_LABELS_2025-01-22_1419.csv", n_max = 1)

# Split the header string on commas, handling quoted strings
label_names <- read.csv(text = header_line, header = FALSE, stringsAsFactors = FALSE) |> 
  as.character()

# Confirm result
label_names

# Make sure they align
stopifnot(length(colnames(redcap_raw)) == length(label_names))

# Create a raw label dictionary
label_dictionary <- tibble(
  variable_name = colnames(redcap_raw),
  label = label_names
)

# Save the dictionary
write_csv(label_dictionary, "data/metadata/label_dictionary_raw.csv")
```

## Filtering

### Row Filtering

Filtering out the rows with "logs_arm_1" and "end_of_study_arm_1", as they are not necessary. We will keep only baseline visit and 6 month followup.

```{r}
# Remove any non-data events: logs, end-of-study summaries
redcap_filtered <- redcap_raw %>%
  filter(!redcap_event_name %in% c("logs_arm_1", "end_of_study_arm_1"))

# Preview changes made
redcap_filtered
```

### Column Filtering

We are going off of the list of necessary columns, per the PHD student in my group.

First, we will define a list of variables to keep.

```{r}
# Define variables to keep based on manual curation
keep_vars <- c(
  "subject_id", "redcap_event_name", "reg_age", "reg_grp",
  "hw_wght", "hw_hght", "dem_sex", "dem_eth",
  "dem_ethwht", "dem_ethblck", "dem_ethasn", "dem_ethmix", "dem_ethoth",
  "dem_ukdiag1", "dem_ukdiag2", "dem_ukdiag3", "dem_ukdiag4", "dem_ukdiag5", "dem_ukdiag6",
  "ectrt_corti1", "ectrt_corti1cls", "ectrt_add1", "ectrt_corti1cls_2", "ectrt_add2", "ectrt_corti1cls_3",
  "ectrt_calci1", "ectrt_cakci1drg", "ectrt_calci2", "ectrt_cakci2drg_2", "ectrt_calci3", "ectrt_cakci3drg_3",
  "ectrt_st", "ectrt_st_no", "ectrt_st_add1",
  "mh_asthma", "mh_rhino", "mh_fdalle", "mh_ctalle", "mh_slpdist",
  "sbfa_twel", "sbfa_twel_det2", "sbfa_arm_flux1", "sbfa_arm_flux2", "sbfa_arm_flux3",
  "sbfa_arm_cur", "sbfa_temp", "sbfa_humidity",
  "sbfa_ph", "sbfa_ph_det", "sbfa_ph1", "sbfa_ph2", "sbfa_ph3",
  "se_easi", "bmsi_adverse_score", "bmsi_capable_score", "bmsi_response_score",
  "btms_bm_score", "btms_bthreat_score", "btms_tscore",
  "vas_score", "skinqol_used", "skinqol_score", "patientreported_quality_of_life_measures_complete",
  "sdc_srbd_scale", "sdc_sleepiness_scale", "sdc_restless_leg_scale",
  "sdc_mtainsc", "sdc_mtainsc_tscore",
  "sdsc_breath", "sdc_breath_tscore", "sdsc_arousal", "sdc_arousal_tscore",
  "sdsc_transit", "sdc_transit_tscore", "sdsc_somnol", "sdc_somnol_tscore",
  "sdsc_hyper", "sdc_hyper_tscore", "sdsc_tscore", "total_t_score",
  "ehi_tscore", "wasi_fsiq4", "gonogo_resp", "simon_presp",
  "cpt_prerrors", "mack_mskips", "mack_iskips"
)
```

Then filter for those variables only.

```{r}
# Subset only the selected variables
redcap_core_vars <- redcap_filtered %>%
  select(all_of(keep_vars))

# View preview
redcap_core_vars
```

Lastly, save as redcap_core_vars.csv

```{r}
# Save cleaned core dataset
write_csv(redcap_core_vars, "data/cleaned/redcap_core_vars.csv")
```

## Cleaning

### Renaming Subject IDs and Event Names

Throughout the rest of the study, the subject IDs are in the format MSXX. We want that to be consistent within this database, so we will change the names from 728-X.

There is also an unnecessary "*arm*\_1" in every event name row, which we will remove.

```{r}
# Clean subject_id and redcap_event_name, save as new object
redcap_cleaned <- redcap_core_vars %>%
  # Convert subject_id like "728-X" → "MSXX"
  mutate(subject_id = str_pad(str_extract(subject_id, "\\d+$"), width = 2, pad = "0"),
         subject_id = paste0("MS", subject_id)) %>%
  # Remove "_arm_1" from redcap_event_name
  mutate(redcap_event_name = str_remove(redcap_event_name, "_arm_1"))

# Preview to ensure changes made
redcap_cleaned
```

### Adding in a Visit ID

```{r}
# Add visit_id (e.g., MS01_V1) based on cleaned event names
redcap_cleaned <- redcap_cleaned %>%
  mutate(
    visit_number = case_when(
      redcap_event_name == "baseline__visit_1" ~ "V1",
      redcap_event_name == "6_month_followup" ~ "V2",
      TRUE ~ NA_character_
    ),
    visit_id = paste0(subject_id, "_", visit_number)
  ) %>%
  relocate(visit_id, .after = redcap_event_name)

# Preview to ensure changes made
redcap_cleaned %>% 
  pull(visit_id) %>% 
  head(10)

# Save cleaned dataset
write_csv(redcap_cleaned, "data/cleaned/redcap_cleaned.csv")
```
