---
title: "Mind & Skin Redcap Database"
author: "Ella Jade Alexander"
format: html
editor: visual
---

# Redcap Database

## Setup

### Setting Base Directory

Assuming that this quarto is in the Redcap/scripts folder, we want to set it to the Redcap folder.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))

```

### Loading Necessary Libraries

```{r}
library(tidyverse)
library(janitor)
library(dplyr)
library(readr)
```

### Loading Raw Data

```{r}

# Read Redcap csv with snake_case
redcap_raw <- read_csv("data/raw/MindSkinVer40_DATA_2025-01-22_1418.csv")

# Preview the raw database
redcap_raw
```

### Metadata: Data Dictionary

Here, we are creating a data dictionary that shows the human-readable, snake_case labels, and variable coding all in one place.

When loading the DATA_LABELS csv into R, the names of the columns are renamed for each to be unique, since there are multiple repeats. So we can't just load it in normally.

Also, for this to work, there have to be equal numbers of columns in both .csv files. The stopifnot function ensures that before continuing.

```{r}
# Extract clean column labels from the LABELS file
label_names <- names(read_csv("data/raw/MindSkinVer40_DATA_LABELS_2025-01-22_1419.csv", col_names = TRUE, n_max = 0))

# Read just the first line (header) of the LABELS CSV as a plain string
header_line <- read_lines("data/raw/MindSkinVer40_DATA_LABELS_2025-01-22_1419.csv", n_max = 1)

# Split the header string on commas, handling quoted strings
label_names <- read.csv(text = header_line, header = FALSE, stringsAsFactors = FALSE) |> 
  as.character()

# Confirm result
label_names

# Make sure they align
stopifnot(length(colnames(redcap_raw)) == length(label_names))

# Create a raw label dictionary
label_dictionary <- tibble(
  variable_name = colnames(redcap_raw),
  label = label_names
)

# Save the dictionary
write_csv(label_dictionary, "data/metadata/label_dictionary_raw.csv")
```

## Filtering

### Row Filtering

Filtering out the rows with "logs_arm_1" and "end_of_study_arm_1", as they are not necessary. We will keep only baseline visit and 6 month followup.

```{r}
# Remove any non-data events: logs, end-of-study summaries
redcap_filtered <- redcap_raw %>%
  filter(!redcap_event_name %in% c("logs_arm_1", "end_of_study_arm_1"))

# Preview changes made
redcap_filtered
```

### Column Filtering

We are going off of the list of necessary columns, per the PHD student in my group.

First, we will define a list of variables to keep.

```{r}
# Define variables to keep based on manual curation
keep_vars <- c(
  "subject_id", "redcap_event_name", "reg_age", "reg_grp",
  "hw_wght", "hw_hght", "dem_sex", "dem_eth",
  "dem_ethwht", "dem_ethblck", "dem_ethasn", "dem_ethmix", "dem_ethoth",
  "dem_ukdiag1", "dem_ukdiag2", "dem_ukdiag3", "dem_ukdiag4", "dem_ukdiag5", "dem_ukdiag6",
  "ectrt_corti1", "ectrt_corti1cls", "ectrt_add1", "ectrt_corti1cls_2", "ectrt_add2", "ectrt_corti1cls_3",
  "ectrt_calci1", "ectrt_cakci1drg", "ectrt_calci2", "ectrt_cakci2drg_2", "ectrt_calci3", "ectrt_cakci3drg_3",
  "ectrt_st", "ectrt_st_add1",
  "mh_asthma", "mh_rhino", "mh_fdalle", "mh_ctalle", "mh_slpdist",
  "sbfa_twel", "sbfa_twel_det2", "sbfa_arm_flux1", "sbfa_arm_flux2", "sbfa_arm_flux3",
  "sbfa_arm_cur", "sbfa_temp", "sbfa_humidity",
  "sbfa_ph", "sbfa_ph_det", "sbfa_ph1", "sbfa_ph2", "sbfa_ph3",
  "se_easi", "bmsi_adverse_score", "bmsi_capable_score", "bmsi_response_score",
  "btms_bm_score", "btms_bthreat_score", "btms_tscore",
  "vas_score", "skinqol_used", "skinqol_score", "patientreported_quality_of_life_measures_complete",
  "sdc_srbd_scale", "sdc_sleepiness_scale", "sdc_restless_leg_scale",
  "sdc_mtainsc", "sdc_mtainsc_tscore",
  "sdsc_breath", "sdc_breath_tscore", "sdsc_arousal", "sdc_arousal_tscore",
  "sdsc_transit", "sdc_transit_tscore", "sdsc_somnol", "sdc_somnol_tscore",
  "sdsc_hyper", "sdc_hyper_tscore", "sdsc_tscore", "total_t_score",
  "ehi_tscore", "wasi_fsiq4", "gonogo_resp", "simon_presp",
  "cpt_prerrors", "mack_mskips", "mack_iskips"
)
```

Then filter for those variables only.

```{r}
# Subset only the selected variables
redcap_core_vars <- redcap_filtered %>%
  select(all_of(keep_vars))

# View preview
redcap_core_vars
```

Lastly, save as redcap_core_vars.csv

```{r}
# Save cleaned core dataset
write_csv(redcap_core_vars, "data/cleaned/redcap_core_vars.csv")
```

## Cleaning

### Labels

```{r}
# Load label dictionary
label_dict <- read_csv("data/metadata/label_dictionary_clean.csv")

# Create a named vector for renaming: names = new (clean), values = current (snake_case)
rename_vector <- setNames(label_dict$snake_case, label_dict$clean)

# Safety check: preview which columns will be renamed
matched_cols <- intersect(names(rename_vector), colnames(redcap_core_vars))
cat("Columns that will be renamed:\n")
print(rename_vector[matched_cols])

# Rename the columns
redcap_clean_labels <- redcap_core_vars %>%
  rename(any_of(rename_vector))

# Preview result
redcap_clean_labels
```

### Renaming Subject IDs and Event Names

Throughout the rest of the study, the subject IDs are in the format MSXX. We want that to be consistent within this database, so we will change the names from 728-X.

There is also an unnecessary "*arm*\_1" in every event name row, which we will remove.

```{r}
# Clean subject_id and redcap_event_name, save as new object
redcap_clean <- redcap_clean_labels %>%
  mutate(
    # Convert subject_id like "728-X" → "MSXX"
    subject_id = str_pad(str_extract(subject_id, "\\d+$"), width = 2, pad = "0"),
    subject_id = paste0("MS", subject_id),

    # Convert visit_number (already renamed) into numeric ID: 1 = baseline, 2 = follow-up
    visit_number = case_when(
      visit_number == "baseline_visit_1" ~ 1L,
      visit_number == "6_month_followup" ~ 2L,
      TRUE ~ NA_integer_
    )
  )

# Preview to ensure changes made
redcap_clean


# Save cleaned dataset
write_csv(redcap_clean, "data/cleaned/redcap_clean.csv")
```

### Further Name Cleaning

## Database Schema

### Core Tables

#### Participant

```{sql}
CREATE TABLE participant (
    participant_id INT PRIMARY KEY AUTO_INCREMENT,
    subject_id VARCHAR(10) NOT NULL,
    age DECIMAL(4,2),
    grp INTEGER,
    sex INTEGER,
    mh_asthma INTEGER,
    mh_rhino INTEGER,
    mh_fdalle INTEGER,
    mh_ctalle INTEGER,
    mh_slpdist INTEGER,
);
```

ethnicity

```         
eth INTEGER,
ethwht INTEGER,
ethblck INTEGER,
ethasn INTEGER,
ethmix INTEGER,
ethoth INTEGER,
```

#### VIsit

```{sql}
CREATE TABLE visit (
    visit_id INT PRIMARY KEY AUTO_INCREMENT,
    visit_number VARCHAR(50) NOT NULL,
    hw_wght DECIMAL(6,2),
    hw_hght DECIMAL(6,2)
    
    FOREIGN KEY (participant_id) REFERENCES participants(subject_id)
);
```

#### UK Diagnostic Criteria

```{sql}
CREATE TABLE uk_diagnostic_criteria (
    ukdiag1 INTEGER,
    ukdiag2 INTEGER,
    ukdiag3 INTEGER,
    ukdiag4 INTEGER,
    ukdiag5 INTEGER,
    ukdiag6 INTEGER
    
    FOREIGN KEY (visit_id) REFERENCES visit(visit_id)
);
```

#### Treatments

```{sql}
CREATE TABLE treatments (
    corti1 INTEGER,
    corti1_potency INTEGER,
    corti2 INTEGER,
    corti2_potency INTEGER,
    corti3 INTEGER,
    corti3_potency INTEGER,
    calci1 INTEGER,
    cakci1_name VARCHAR(50),
    calci2 INTEGER,
    cakci2_name VARCHAR(50),
    calci3 INTEGER,
    calci3_potency VARCHAR(50),
    ectrt_st INTEGER,
    ectrt_st_no INTEGER,
    ectrt_st_add1 INTEGER,
    
    FOREIGN KEY (visit_id) REFERENCES visits(visit_id)
);
```

OR

```{sql}

CREATE TABLE treatments (
    treatment_id INTEGER PRIMARY KEY AUTO_INCREMENT,
    visit_id VARCHAR(20) NOT NULL,
    treatment_type VARCHAR(20) NOT NULL, -- 'corticosteroid', 'calcineurin', 'steroid'
    treatment_number INTEGER NOT NULL, -- 1, 2, 3 for multiple treatments
    strength_class INTEGER
    
    FOREIGN KEY (visit_id) REFERENCES visit(visit_id)
);
```

#### Skin Assessments

```{sql}
CREATE TABLE skin_assessments (
    skin_id VARCHAR(20) PRIMARY KEY,
    sbfa_twel INTEGER,
    sbfa_twel_det2 INTEGER,
    sbfa_arm_flux1 DECIMAL(8,3),
    sbfa_arm_flux2 DECIMAL(8,3),
    sbfa_arm_flux3 DECIMAL(8,3),
    sbfa_arm_cur INTEGER,
    sbfa_temp DECIMAL(5,2),
    sbfa_humidity DECIMAL(5,2),
    sbfa_ph INTEGER,
    sbfa_ph_det INTEGER,
    sbfa_ph1 DECIMAL(4,2),
    sbfa_ph2 DECIMAL(4,2),
    sbfa_ph3 DECIMAL(4,2),
    
    FOREIGN KEY (visit_id) REFERENCES visits(visit_id)
);
```

#### Clinical Severity Scores

```{sql}
CREATE TABLE clinical_severity_scores (
    severity_id VARCHAR(20) PRIMARY KEY,
    se_easi DECIMAL(5,2),
    vas_score INTEGER,
    skinqol_used INTEGER,
    skinqol_score INTEGER,
    
    FOREIGN KEY (visit_id) REFERENCES visits(visit_id)
);
```

#### Psychological Assessments

```{sql}
CREATE TABLE psychological_assessments (
    psych_id VARCHAR(20) PRIMARY KEY,
    bmsi_adverse_score DECIMAL(5,2),
    bmsi_capable_score INTEGER,
    bmsi_response_score DECIMAL(5,2),
    btms_bm_score INTEGER,
    btms_bthreat_score INTEGER,
    btms_tscore INTEGER,
    
    FOREIGN KEY (visit_id) REFERENCES visits(visit_id)
);
```

#### Sleep Assessments

```{sql}
CREATE TABLE sleep_assessments (
    sleep_id VARCHAR(20) PRIMARY KEY,
    sdc_srbd_scale DECIMAL(5,3),
    sdc_sleepiness_scale DECIMAL(4,3),
    sdc_restless_leg_scale DECIMAL(4,3),
    sdc_mtainsc INTEGER,
    sdc_mtainsc_tscore INTEGER,
    sdsc_breath INTEGER,
    sdc_breath_tscore INTEGER,
    sdsc_arousal INTEGER,
    sdc_arousal_tscore INTEGER,
    sdsc_transit INTEGER,
    sdc_transit_tscore INTEGER,
    sdsc_somnol INTEGER,
    sdc_somnol_tscore INTEGER,
    sdsc_hyper INTEGER,
    sdc_hyper_tscore INTEGER,
    sdsc_tscore INTEGER,
    total_t_score INTEGER,
    
    FOREIGN KEY (visit_id) REFERENCES visits(visit_id)
);
```

#### Cognitive Assessments

```{sql}

CREATE TABLE cognitive_assessments (
    cognitive_id VARCHAR(20) PRIMARY KEY,
    ehi_tscore DECIMAL(6,2),
    wasi_fsiq4 INTEGER,
    gonogo_resp INTEGER,
    simon_presp INTEGER,
    cpt_prerrors INTEGER,
    mack_mskips INTEGER,
    mack_iskips INTEGER,
    
    FOREIGN KEY (visit_id) REFERENCES visits(visit_id)
);
```
