---
title: "Dreem Analysis"
author: "Ella Jade Alexander"
format: html
editor: visual
---

# Mind&Skin Dreem Analysis

## Overview

### Analysis Goals

#### 1. **Power Spectral Density (PSD) Comparison Across Groups**

-   **By sleep stage**: For each stage (W, N1, N2, N3, R), compare the PSD.

-   **By group**: Plot PSD for the 3 participant groups (e.g., control, mild, severe).

-   **Channels**: Use **F7 and F8** (especially: EEG_F8_O2, EEG_F7_O2).

-   **Plot type**: Frequency (Hz) on x-axis, Power on y-axis. One line per group.

#### 2. **Statistical Comparison**

-   Extract total power in specific bands (e.g., delta: 0.5–4 Hz).

-   Run statistical comparisons (1-way ANOVA) across the 3 groups.

-   Use per-subject averaged values from sigstats or PSD.

#### 3. **Variability Over Multiple Nights**

-   For participants with \>1 night:

    -   Measure **intra-subject PSD variability**

    -   Assess device consistency

### Power Spectral Density (PSD) Analysis by Sleep Stage and Group

#### What is PSD?

Power Spectral Density (PSD) is a method used to quantify the distribution of power across different frequency bands in EEG signals. It transforms time-domain data (e.g., EEG voltage over time) into frequency-domain representations, allowing us to evaluate how much power (i.e., signal strength) exists in specific frequency ranges, such as delta (0.5–4 Hz), theta (4–8 Hz), alpha (8–12 Hz), and others.

In the context of sleep EEG, different frequency bands are known to correlate with different physiological and cognitive states. For example:

-   **Delta power (0.5–4 Hz)** is prominent in deep (N3) sleep.

-   **Alpha power** is typically reduced in sleep but can be a marker of wakefulness or arousal.

-   **Spindle and beta activity** may relate to sleep architecture stability or disturbance.

#### Why PSD is Useful for This Study

In this pediatric eczema study, comparing PSD across severity groups allows us to:

-   **Quantify sleep differences** objectively beyond clinical scores or subjective reports.

-   **Detect neurophysiological signatures** of inflammation, itching, or sleep fragmentation.

-   **Compare signal characteristics** across groups in each sleep stage to see if certain EEG dynamics are altered in more severe eczema.

This analysis is especially relevant given the Dreem device’s multi-night home-based acquisition — allowing assessment of **naturalistic, ecologically valid sleep data** across groups.

#### Goals of This Step

1.  **Compare PSD across sleep stages and groups**

    -   For each sleep stage (W, N1, N2, N3, R), we will generate PSD curves.

    -   Each curve will show average PSD (power vs. frequency) for each group.

    -   One plot per stage, with 3 lines (Group 1–3).

    -   Frequency (Hz) on x-axis, power (uV²/Hz) on y-axis.

2.  **Use clean and consistent EEG signals**

    -   Only channels involving F7 and F8 (frontal) will be analyzed.

    -   These channels were chosen due to lower noise in Dreem recordings, per prior QC and Luna guidance.

    -   Specific channels include: `EEG_F7_O1,EEG_F8_O2,EEG_F8_F7,EEG_F8_O1,EEG_F7_O2`

3.  **Extract summary statistics from PSD**

    -   Using Luna’s `SIGSTATS` command, we can compute total power within specific frequency bands.

    -   For example, we will extract **delta power** (0.5–4 Hz) and compare across groups.

    -   This becomes input for statistical tests such as **one-way ANOVA**.

4.  **Assess night-to-night variability**

    -   For participants with multiple nights, we will compare PSD estimates across those nights.

    -   This gives insight into:

        -   **Intra-subject variability**

        -   **Signal stability across nights**

        -   **Device consistency**

#### Why Stratifying by Sleep Stage Matters

Each sleep stage has unique EEG spectral characteristics:

-   **Wake (W)**: Higher alpha/beta

-   **N1**: Mixed low-frequency patterns

-   **N2**: Spindles, theta

-   **N3**: Dominated by delta power

-   **REM (R)**: Lower amplitude, mixed frequencies

Analyzing PSD without stratifying by stage would collapse meaningful variation — so Luna’s ability to compute stage-specific PSD (`PSD STAGE`) is crucial.

#### Tools Used

We use Luna’s:

-   `PSD` command to estimate spectral power

-   `SIGSTATS` to extract band-specific power summaries

-   `CHEP` to apply signal quality masks (prior step)

-   `destrat` to extract outputs for R

-   R to merge PSD data with `id_group_map.csv` for visualization and statistical testing

## Setup

### Load Necessary Libraries

```{r}
library(luna)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(readr)
library(stringr)
library(lubridate) # for working with dates
library(scales)   # to access breaks/formatting functions
library(gridExtra)
library(tidyr)
```

### Load the Database

```{r}
# Load in the Database
chep_psd_sigstats <- ldb("../../data/processed/luna_outputs/dreem_chep_psd_sigstats.db")
```

```{r}
chep_psd_sigstats_stage <- ldb("../../data/processed/luna_outputs/dreem_chep_psd_sigstats_stage.db")
```

```{bash}
# W stage
luna /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/sample_list/min_4hr_edf.lst \
  -o /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/luna_outputs/dreem_psd_by_stage.db \
  -s 'EPOCH & MASK ifnot=W & RE & TAG STAGE/W & PSD spectrum'

# N1 stage
luna /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/sample_list/min_4hr_edf.lst \
  -a /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/luna_outputs/dreem_psd_by_stage.db \
  -s 'EPOCH & MASK ifnot=N1 & RE & TAG SS/N1 & PSD spectrum'

# N2 stage
luna /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/sample_list/min_4hr_edf.lst \
  -a /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/luna_outputs/dreem_psd_by_stage.db \
  -s 'EPOCH & MASK ifnot=N2 & RE & TAG SS/N2 & PSD spectrum'

# N3 stage
luna /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/sample_list/min_4hr_edf.lst \
  -a /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/luna_outputs/dreem_psd_by_stage.db \
  -s 'EPOCH & MASK ifnot=N3 & RE & TAG SS/N3 & PSD spectrum'

# REM stage
luna /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/sample_list/min_4hr_edf.lst \
  -a /Users/ellajadealex/Desktop/Mind_Skin_Dissertation/Dreem/data/processed/luna_outputs/dreem_psd_by_stage.db \
  -s 'EPOCH & MASK ifnot=R & RE & TAG SS/R & PSD spectrum'

```

```{r}
dreem_psd_by_stage <- ldb("../../data/processed/luna_outputs/dreem_psd_by_stage.db")
```

Current issues:\

## Analysis

### Step 1: Extract PSD and SIGSTATS Tables

Luna’s output database (`dreem_chep_psd_sigstats.db`) already contains structured tables produced from prior analysis steps using the `PSD` and `SIGSTATS` commands. When loaded with `ldb()`, the database is parsed into a list of data frames that reflect different levels of stratification (e.g., per-channel, per-epoch, per-frequency).

There is no need to use `destrat()` here because the database format is already optimized for direct access.

We extract two key tables:

-   `PSD$CH_F`: contains power values (`PSD`) at each frequency (`F`) for each channel and session.

-   `SIGSTATS$CH_E`: contains per-epoch band power summaries (`H1`, `H2`, etc.) for each channel and sleep stage.

These are the primary inputs for PSD visualization and summary statistic extraction.

```{r}
# Extract per-frequency PSD (for spectral curves)
psd <- chep_psd_sigstats$PSD$CH_F

# Extract per-epoch SIGSTATS (for band power analysis)
sig <- chep_psd_sigstats$SIGSTATS$CH_E
```

## Step 2: Merge PSD/SIGSTATS with Group Metadata

To compare spectral features across participant groups (e.g., control, mild, severe eczema), we must first link each session’s `standard_id` to its assigned group. We previously created this mapping in `id_group_map.csv`.

This step ensures that both the `psd` and `sig` tables include a `group` column — which is necessary for plotting group-level averages and performing statistical comparisons.

We do this by:

1.  Reading the metadata file (`id_group_map.csv`) containing the `standard_id` and `group` columns.

2.  Renaming the `ID` column in both `psd` and `sig` tables to `standard_id` to match the metadata.

3.  Performing a `left_join()` to add the `group` to each row.

```{r}
# Load group assignment metadata
groups <- readr::read_csv("../../data/processed/metadata/id_group_map.csv")

# Join group metadata to PSD data
psd <- psd %>%
  rename(standard_id = ID) %>%
  left_join(groups, by = "standard_id")

# Join group metadata to SIGSTATS data
sig <- sig %>%
  rename(standard_id = ID) %>%
  left_join(groups, by = "standard_id")
```

## Step 3: Filter PSD by Sleep Stage and EEG Channel

To visualize group differences, we start by selecting **one combination** of:

-   **Sleep stage** (e.g., N3 — deep sleep)

-   note: for now we are using all channels but that can be changed later

This lets us generate a clean line plot of **power vs. frequency**, averaged within each group.

We’ll begin with:

-   **Stage:** `N3`

Later, we can scale this to loop or facet over more combinations.

```{r}
names(chep_psd_sigstats)
names(chep_psd_sigstats$PSD)
```

```{r}
psd_n3 <- psd %>%
  filter(STAGE == "N3")
```
